Glyph-maps for visualizing climate data and models

### 2011-08-26 paper.tex
Fleshing out computation. Move to new location.  Add more references

%An interesting attribute of these glyph plots is that the position of graphical elements is determined by major and minor axes. For maps, the major axes are latitude and longitude, and the minor axes are typically time on the x-axis and some measurement or model prediction on the y-axis.

@@

%Will first explore for regularly gridded data, a simple starting point and a common output of climate models, and then show a few simple extensions to deal with non-rectangular grids and irregular locations, as exemplified by raw climate data which is typically collect at irregular locations.
\section{Exploring trend}

\begin{figure*}[htp]
\centerline{\includegraphics[width=4in]{nasa-deseas-trend.pdf}}
\caption{Time series icon at each location, showing de-seasonalized trend.}
\end{figure*}

\section{Showing seasonal effect, radially oriented icon}

A big advantage of glyph-maps is that they can be generated with existing graphics software, after a simple pre-processing step. This makes them immediately accessible to a wide audience. The graphics in this paper were created in R \citep{R} using ggplot2 \citep{me:ggplot2,wickham:2007d} and \citep{me:plyr}, but are readily implemented in any statistical programming environment.  Code and data is available in the online supplementary materials.

Because the x and y locations are linear combinations of two variables, these types of displays can also be built interactively by software that supports tours, such as DataViewer, XGobi or GGobi.  This technique is used to good effect in \citet{buja:1996a}.

@@

\section{Radial icons/star glyphs}

These time series glyphs can be readily altered to support circular variables, by projecting each line into polar coordinates and joining the ends. This type of display is best suited for variables that are truly circular, such as seasonal patterns or angles (such as for wind direction).

Cite star glyphs.

The alteration to the data transformation is quite simple. Without loss of generality, we assume that $y_{minor}$ becomes $r$, the radius, and $x_{minor}$ becomes $theta$ the angle, both rescaled to lie between 0 and 1. Then the glyph $x$ coordinate is given by $x_{major} + \frac{w \cdot r}{2} \cos(2 \pi \theta)$, and the $y$ coordinate by $y_{major} + \frac{h \cdot r}{2} \sin(2 \pi \theta)$.

>>>>

@@

%Will first explore for regularly gridded data, a simple starting point and a common output of climate models, and then show a few simple extensions to deal with non-rectangular grids and irregular locations, as exemplified by raw climate data which is typically collect at irregular locations.

Code and data is available in the online supplementary materials.

\section{Computation}

A big advantage of glyph-maps is that they can be generated with existing graphics software after performing a simple pre-processing step described in this section. This makes them immediately accessible to a wide audience. The graphics in this paper were created in R \citep{R} using ggplot2 \citep{me:ggplot2,wickham:2007d} and \citep{me:plyr}, but glyph-maps are readily implemented in any statistical programming environment.

The pre-processing step is simple, once you notice that the position of graphical elements is determined by a major and minor axis in each direction (x and y). For maps, the major axes are latitude and longitude, and the minor axes are typically time on the x-axis and some measurement or model prediction on the y-axis. The final locations are a simple linear combination.  After rescaling the minor axis to have range $[0, 1]$, the $x$ coordinate is given by $x_{major} + w \cdot x_{minor}$ and $y$ by $y_{major} + h \cdot y_{minor}$, where $w$ and $h$ are width and height respectively

Because the x and y locations are linear combinations of two variables, these types of displays can also be built interactively by software that supports tours \citep{cook:2006}, such as DataViewer \citep{buja:1986}, XGobi \citep{swayne:1991} or GGobi \citep{swayne:2003}.  This technique is used to good effect in \citet{buja:1996a}.

Computing the coordinates for star glyphs is equally easy. Without loss of generality, we assume that $y_{minor}$ becomes $r$, the radius, and $x_{minor}$ becomes $\theta$, the angle, both rescaled to lie between 0 and 1. Then the glyph $x$ coordinate is given by $x_{major} + \frac{w \cdot r}{2} \cos(2 \pi \theta)$, and the $y$ coordinate by $y_{major} + \frac{h \cdot r}{2} \sin(2 \pi \theta)$.

Figure~\ref{fig:cycle} shows monthly temperature effects as time (Cartesian coordinates) and star (polar coordinates) glyphs. In this case, the star glyphs are not as effective as the linear glyphs: perhaps because of the absence of cyclical pattern.

@@
